P3692R0: How to Avoid OOTA Without Really Trying

### 2025-06-11 paper/ootavacuous.tex
More minor updates
More minor updates to the text, mostly concerned with the introductions
of the notions of semantic and syntactic dependencies.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly recommends against them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no ``out-of-thin-air'' values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.
(The as-if rule permits optimizations where the optimized code's
observable behavior is the same as if it were unoptimized.)

@@

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
T0 and T1.
Roughly speaking, there is a semantic dependency from a given load to
a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
(A \emph{syntactic} dependency, by contrast, exists whenever the value
of the load appears in the source code as part of the computation
leading up to the store, whether it affects the store or not.
Only semantic dependencies are relevant to OOTA.)
Here the semantic dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.

@@

The key constraints mentioned above are:
(1)~Communicating data within a computer system takes time, and
(2)~Executing instructions that compute an output value that semantically
depends on given input values also takes time.
Because each link in an OOTA cycle consists of either data communication
or computation of an output that semantically depends on an input,
and because closing an OOTA cycle comprising a sequence of commmunications
and computations requires this sequence to end at the same time as it
starts, OOTA cycles cannot form in those C++ implementations that are
within the scope of this paper.
A great advantage of our approach is that we do \emph{not} need to define
semantic dependency purely at the level of C++ source code;
rather, we are able to utilize information already provided by existing
hardware execution models.

These simple-seeming ideas conceal considerable complexity,
and much of the remainder of this paper deals with corner cases that
arise when carefully considering semantic dependencies, executions, C++
compilers, and hardware properties.

>>>>

Because nothing in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly recommends against them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no ``out-of-thin-air'' values
	are computed that circularly depend on their own computation.
\end{quote}
This prohibition prevents misapplication of the as-if rule in oracular
C++ implementations.
(The as-if rule permits optimizations for which the optimized code's
observable behavior is the same as if it were unoptimized.)

@@

Unfortunately, the standard does not explain what the phrase ``depend on''
in the quotation above really means.
It is generally taken to mean \emph{semantic dependency},
a somewhat vague concept.
Roughly speaking, there is a semantic dependency from a given load to
a given store in the same thread when
\emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
In Listing~\ref{lst:Simple OOTA}, for example,
there is a semantic dependency from line~1 to line~2 in both T0 and T1.
The semantic dependencies here are trivial,
because the values stored simply \emph{are} the values that were loaded.
(A \emph{syntactic} dependency, by contrast, exists whenever the value
of the load appears in the source code as part of the computation
leading up to the store, whether it affects the store or not.
\emph{@@@ Reference?}
Only semantic dependencies contribute to OOTA.)

@@

The key constraints mentioned above are:
(1)~Communicating data within a computer system takes time, and
(2)~Executing instructions that compute an output value that semantically
depends on given input values also takes time.
Because each link in an OOTA cycle consists of either data communication
or computation of an output that semantically depends on an input,
and because closing such a cycle requires the sequence of links
to end at the same time as it
starts, OOTA cycles cannot form in those C++ implementations that are
within the scope of this paper.
A great advantage of our approach is that we do \emph{not} need to define
semantic dependency purely at the level of C++ source code;
rather, we are able to utilize information already provided by existing
hardware execution models.

These simple-seeming notions conceal considerable complexity,
and much of the remainder of this paper deals with corner cases that
arise when carefully considering semantic dependencies, executions, C++
compilers, and hardware properties.
